---
title: "P.berghei pir analysis"
author: "Timothy Little"
date: "31/08/2021"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, results = TRUE, message = FALSE, warning = FALSE)
options(scipen = 999) # Means that numbers are displayed normally not as 1e6 (1000000)
```

```{r load_packages, include = FALSE, message = FALSE}
library('readxl')
library('readr')
library('dplyr')
library('purrr')
library('ggplot2')
library("RColorBrewer")
library('colorspace')
library('data.table')
library('gridExtra')
library('tibble')
library('scales')
library('stringr')
library('viridis')
library('corrplot')
library('reshape2')
library('circlize') # for circular plots
library('plotly')
library('ComplexHeatmap')

#Colour blind friendly palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
show_col(cbPalette)

#Want the defaults for arrange, select, mutate and summarise to be from dplyr
arrange <- dplyr::arrange
mutate <- dplyr::mutate
summarise <- dplyr::summarise
select <- dplyr::select

```

```{r sourcing_functions}
source('../rfunctions/uniquify_fun.R')
source('../rfunctions/ordering_fun.R')
source('../rfunctions/tpm_fun.R')
source('../rfunctions/usefulFunctions_mround_my-max_firstup_fun.R')
```

```{r ordering_vectors}
order_vector <- c('Sporo', 'Liv-pre4h', 'Liv-pre47h', 'Liv-post48h',
                 'Liv-2h','Liv-4h','Liv-12h', 'Liv-18h',
                 'Liv-24h' ,'Liv-36h','48h','54h','60h',
                 'Mero','Asex','Schiz','Gam','-Pre3h','BldMl-24h','Ook')

r_order_vector <- str_replace_all(order_vector, 
                                  pattern = '-', 
                                  replacement = '\\.')
```

```{r pir_id, message=FALSE, results='hide'}

# Getting the pir gene names and transcript lengths needed for calculating the transcript-per-million

# First we need to get the _pir gene names. These have been downloaded using a search of PlasmoDB P. berghei ANKA genome v3 using  "\*pir\*" search and manual removal of genes which clearly were not pir.

allgene.info.file <- '../standard_files/PlasmoDb-47_beta_Pberghei_GenesByTaxon_Summary.csv'
allgene.info <- as_tibble(
  read_csv(allgene.info.file)
) %>%
  mutate(Geneid = `Gene ID`, 
         Description = `Product Description`,
         is_pseudo = `Is Pseudo`,
         transcript_length = `Transcript Length`,
         paralog_count = `Paralog count`) %>% 
  select(Geneid, 
         Description, 
         is_pseudo, 
         transcript_length, 
         paralog_count)

pir_info <- allgene.info %>% 
  filter(str_detect(Description, 
                    pattern = 'PIR(?!.+pseud)'), 
         is_pseudo == 'No') %>% 
  filter(!Geneid %in% 'PBANKA_0524600')
#Removing the pir-like gene.

pir_id <- pir_info$Geneid

#pir

pir_genes_list <- lapply(
  unique(pir_info$Description), 
  function(pir_group){
    select(
      filter(
        pir_info, 
        Description %in% pir_group
      ),
      Geneid
    )
  }
)
names(pir_genes_list) <- unique(pir_info$Description)

#How many pirs are present on each chromosome?

pir_tbl = as_tibble(data.frame(Geneid = pir_id))
pir_chrom_tbl = pir_tbl %>% 
  mutate(chrom = unlist( #Getting the chromosomes by a complicated set of lapply and sapplys...
    sapply(
      lapply(
        sapply(
          as.character(pir_tbl$Geneid), 
          strsplit,
          split = ''
        ), 
        '[', 
        8:9
      ), 
      paste, 
      collapse = ''
    )
  )
  ) 
pir_chrom_tbl %>% 
  group_by(chrom) %>% 
  dplyr::summarise(number_of_pirs_in_each_chromosome = n(),
                   percentage_of_all_pirs = round(n()/nrow(pir_tbl), 3)*100) %>% 
  dplyr::arrange(desc(percentage_of_all_pirs))

#Getting the loci information
pir_loci <- data.frame(
  Geneid = pir_id,
  chrom = str_extract(pir_id, 
                      pattern = '(?<=_)[[:digit:]]{2}'),
  number = as.numeric(str_extract(pir_id, 
                                  pattern = '(?<=_[[:digit:]]{2})[[:digit:]]+'))
) %>% 
  filter(!chrom %in% '00') %>% #Remove chrom00 (contigs)
  group_by(chrom) %>% 
  mutate(jump = ifelse(lag(number, 
                           default = 1001) < 1000 & number - lag(number, 
                                                                 default = 0) > 1000,
                       1,
                       0)) %>% 
  mutate(
    jump = ifelse(cumsum(jump) == 1 | jump == 1, 1, 0)
  ) %>% 
  mutate(
    locus = paste0(
      chrom,
      ifelse(jump == 0 & number < 1000, 
             'Left',
             'Right')
    )
  )

# Detect 'jumps' of the gene number over 1000 within each chromosome, setting the lag default as 1001 in the
#   first part of the ifelse statement so that chrom2 doesn't have an erroneous jump as it starts at >1000
# Later cumsum is do that every gene after the 'jump' is included allowing us to define telomere ends.

pir_clade <- read_delim('../standard_files/Pberghei_pir_clades.txt', 
                        delim = '\t')
pir_clade <- mutate(pir_clade, 
                    locus = pir_loci$locus[match(Geneid, pir_loci$Geneid)])
pir_id <- pir_clade$Geneid

pir_clade_list <- lapply(unique(pir_clade$clade), function(clade_name){
  pir_clade$Geneid[pir_clade$clade %in% clade_name]
})
names(pir_clade_list) <- as.character(unique(pir_clade$clade))
```

## Reading the TPM 

```{r read_tpm}
berghei_tpm <- read.csv('berghei_tpm.csv', 
                        row.names = 1)
```

Combining the data across the experiments.

```{r combining_stages_data}
combined_data <- berghei_tpm %>% 
  melt(value.name = 'tpm') %>% 
  mutate(stage = str_extract(variable, 
                             pattern = '(?<=_).+(?=_)')) %>% 
  group_by(Geneid, stage) %>% 
  dplyr::summarise(tpm_mean = mean(tpm),
                   tpm_med = median(tpm),
                   tpm_sd = sd(tpm),
                   tpm_max = max(tpm),
                   tpm_min = min(tpm))


combined_data <- dcast(combined_data, 
                       Geneid ~ stage, 
                       value.var = 'tpm_mean') %>% 
  column_to_rownames(var = 'Geneid')

#Ordering by life cycle progression
combined_data <- combined_data[,
                               ordering(colnames(combined_data), 
                                        order_vector = gsub(order_vector,
                                                            pattern = '-',
                                                            replacement = '\\.')
                               )]

#Check that they still sum to 1e6

apply(combined_data, 2, sum)

```

```{r colours_clades_phase}
SorL_col <- setNames(
  c(    '#F0E442',
        '#56B4E9',
        '#E69F00'),
  sort(
    unique(pir_clade %>% 
             mutate(
               SorL = str_extract(clade, 
                                  pattern = '(L|S(?=[[:digit:]]))|ancestral|pan-rodent')
             ) %>% 
             .$SorL)
  )
)

SorL_col_num <- SorL_col
names(SorL_col_num) <- sapply(names(SorL_col), 
                              function(clade_name) {
                                paste0(
                                  clade_name, 
                                  ' (n = ',
                                  sum(pir_clade$SorL == clade_name),
                                  ')'
                                )
                              })

clade_col <- setNames(
  c('#F0E442',
    '#56B4E9',
    '#009E73',
    '#0072B2',
    '#E69F00',
    '#D55E00',
    '#CC79A7',
    '#999999',
    'black',
    'red'),
  sort(
   unique(pir_clade$clade)
  )
)

clade_col_num <- clade_col
names(clade_col_num) <- sapply(names(clade_col), 
                               function(clade_name) {
                                 paste0(
                                   clade_name, 
                                   ' (n = ',
                                   sum(pir_clade$clade == clade_name),
                                   ')'
                                 )
                               })
```

# pir expression, S or L-form, stacked bar chart, stack for each gene

```{r SorL_stacked_eachGene_fig2a}
  tpm_data <- combined_data %>% 
    rownames_to_column(var = 'Geneid') %>% 
    filter(Geneid %in% pir_id) %>% 
    melt(id.vars = 'Geneid', 
         variable.name = 'stage', 
         value.name = 'tpm') %>% 
  mutate(clade = pir_clade$clade[match(Geneid, 
                                       pir_clade$Geneid)]
         ) %>% 
  group_by(clade) %>% 
  mutate(clade = paste0(clade, ' (n = ', n_distinct(Geneid) ,')')) %>% 
    mutate(stack_group = Geneid) %>% 
    mutate(tpm = ifelse(tpm > 1, tpm, 0)) %>% #Filtering step for low value pir genes.
    group_by(stage, 
             stack_group,
             clade) %>% 
    dplyr::summarise(total_tpm = sum(tpm)) %>%
  arrange(
          desc(clade),
          total_tpm
)
  
tpm_data <- mutate(tpm_data,
                   order = row_number())

tpm_data_SorL_stackgenes <- mutate(tpm_data,
                        SorL = str_extract(clade, 
                                           pattern = '(L|S(?=[[:digit:]]))|ancestral|pan-rodent'))
tpm_data_SorL_stackgenes <- group_by(tpm_data_SorL_stackgenes,
                                     stage, 
                                     SorL,
                                     stack_group) %>% 
  dplyr::summarise(total_tpm = sum(total_tpm)) %>% 
  #Include the number of genes in each fill category for the labels
  mutate(SorL = paste0(SorL, 
                       ' (n = ', 
                       n_distinct(stack_group),
                       ')')) %>% 
#Order by SorL and then by total_tpm, then convert stack_group to a factor so ggplot orders correctly
  arrange(SorL, 
          total_tpm) %>% 
  mutate(stage_num = as.integer(stage))
tpm_data_SorL_stackgenes$stage %>% as.integer %>% unique
tpm_data_SorL_stackgenes$stage %>% unique

tpm_data_SorL_stackgenes$stack_group_order <- seq_len(length.out = nrow(tpm_data_SorL_stackgenes))
```

```{r stacked_bar_chart_figure2b}
#Calculate the range for the y value to make it look nicer
stacked_max <- tpm_data %>% 
  ungroup %>% 
  group_by(stage) %>% 
  dplyr::summarise(max_tpm = sum(total_tpm)) %>% 
  .$max_tpm %>% 
  max

max_tpm <- c(
  seq(
    0,
    mround(
      stacked_max,
      500),
    length.out = 11)
)

stack_label <- 'sub-family'

write_csv(x = tpm_data_SorL_stackgenes,
          path = 'tl20-05_SorL-subfamily_pir.csv')

stage_vec <- unique(
  levels(tpm_data$stage)
)

stage_vec <- tpm_data_SorL_stackgenes$stage
col_group <- case_when(str_detect(stage_vec, pattern = 'Asex') ~ 'zzAsex' ,
                       str_detect(stage_vec, pattern = 'Liv') ~ 'zzLiv',
                       str_detect(stage_vec, pattern = 'Gam') ~ 'zzGam',
                       str_detect(stage_vec, pattern  = 'Bld|Sporo|Ook') ~ 'zzMos')
tpm_data_SorL_stackgenes$stage_col <- col_group
fills <- c('#F6B9AD','#96d3ff','#f0d4e3','#a7e9af') #Note that this order is influenced by sort(col_group)
```

```{r draw_stacked_bar_chart}
# Note that this code does not work with ggplot 3.3.5, but it does work with 3.3.3. 
#  Some issue with the colours for the geom_rect background.

y <- ggplot(data = tpm_data_SorL_stackgenes, 
            aes(
              x = stage_num,
              y = total_tpm, 
              fill = SorL,
              # group = stage_num
            )) +
  geom_rect(
    aes(
      xmin = stage_num - .7,
      xmax = stage_num + .7,
      ymin = -Inf,
      ymax = Inf,
      fill = stage_col
    ),
    alpha = .05
  ) +
  geom_bar(position = position_stack(),
           stat = 'identity',
           colour = 'black',
           size = 0.1) + 
  scale_fill_manual(
    breaks = unique(tpm_data_SorL_stackgenes$SorL),
    values = c(
      SorL_col_num,
      fills
    )
  ) +
  scale_color_manual('black') + 
  # labs(fill = stack_label) +
  scale_y_continuous(breaks = max_tpm) +
  theme_classic() +
  # scale_size(range = c(0,3)) +
  ylab(
    bquote('total '~italic(.('pir'))~' TPM')
  ) +
  scale_x_continuous(
    breaks = tpm_data_SorL_stackgenes$stage_num %>% unique(),
    labels = tpm_data_SorL_stackgenes$stage %>% unique(),
    expand = c(0, 0)
  ) +
    theme(
      axis.text.x = element_text(angle = 90, 
                                 vjust = .5,
                                 size = 10),
      axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.key.size = unit(3,'mm'),
    legend.key.width = unit(2,'mm'),
    ) +
    xlab('stage') + 
    facet_grid(.~stage, 
               scales = "free_x",
               space = "free_x") +
    scale_x_discrete(expand=c(0,0)) +
    theme(
      axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x = element_blank(),
      # axis.line.x = element_blank(),
      axis.text.y = element_text(size = 10),
      axis.title.y = element_text(size = 10),
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 10),
      legend.key.size = unit(3,'mm'),
      legend.key.width = unit(2,'mm'),
      strip.text.x = element_text(size=8, 
                                  face = 'bold', 
                                  angle = 90),
      strip.background = element_rect(
        color="black", 
        size=0.1, 
        linetype="solid"
      ),
      panel.spacing.x = unit(0, "null")
    ) 
y
```

```{r change_facet_cols}

g <- ggplot_gtable(ggplot_build(y))
strip_t <- which(grepl('strip-t', g$layout$name))
stage_vec <- unique(
  levels(tpm_data$stage)
)
fills <- case_when(str_detect(stage_vec, pattern = 'Asex') ~ 'coral1' ,
                   str_detect(stage_vec, pattern = 'Liv') ~ 'plum2',
                   str_detect(stage_vec, pattern = 'Gam') ~ 'lightskyblue1',
                   str_detect(stage_vec, pattern  = 'Bld|Sporo|Ook') ~ 'lightgreen')
k <- 1
for (i in strip_t) {
  j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
  g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
  k <- k+1
}

grid.newpage()
grid.draw(g)

```

# figure2binset

```{r stack_bar_graph_SorL_lowexpression-figure2binset}
#Include all stages expressed lower than a certain amount

low_express <- tpm_data_SorL_stackgenes %>% 
  group_by(stage) %>% 
  summarise(sum_tpm = sum(total_tpm)) %>% 
  filter(sum_tpm < 350)

#Calculate the range for the y value to make it look nicer

stacked_max <- max(low_express$sum_tpm)

max_tpm <- c(
  seq(
    0,
    mround(
      stacked_max,
      100),
    length.out = 11)
)

tpm_data_SorL_stackgenes_low <- filter(tpm_data_SorL_stackgenes,
                                       stage %in% low_express$stage)

y <- ggplot(data = tpm_data_SorL_stackgenes_low, 
            aes(
              x = stage,
              y = total_tpm, 
              fill = SorL,
              # group = stage_num
            )) +
  geom_bar(position = position_stack(),
           stat = 'identity',
           colour = 'black',
           size = 0.1) + 
  scale_fill_manual(
    breaks = unique(tpm_data_SorL_stackgenes$SorL),
    values = c(
      SorL_col_num,
      fills
    )
  ) +
  scale_color_manual('black') + 
  labs(fill = stack_label) +
  scale_y_continuous(breaks = max_tpm) +
  theme_classic() +
  scale_size(range = c(0,3)) +
  ylab(
    bquote('total '~italic(.('pir'))~' TPM')
  ) + 
  facet_grid(.~stage, 
             scales = "free_x",
             space = "free_x") +
  scale_x_discrete(expand=expand_scale(add=1)) +
  theme(
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x = element_blank(),
    # axis.line.x = element_blank(),
    axis.text.y = element_text(size = 15),
    axis.title.y = element_text(size = 8),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 5),
    legend.key.size = unit(3,'mm'),
    legend.key.width = unit(2,'mm'),
    strip.text.x = element_text(size=6, 
                                face = 'bold', 
                                angle = 90),
    strip.background = element_rect(
      color="black", 
      size=0.1, 
      linetype="solid"
    ),
    panel.spacing.x = unit(0, "null")
  ) 

#This bit is to change the colour of the facet boxes

g <- ggplot_gtable(ggplot_build(y))
strip_t <- which(grepl('strip-t', g$layout$name))
stage_vec <- unique(
  levels(tpm_data_SorL_stackgenes_low$stage)[levels(tpm_data_SorL_stackgenes_low$stage) %in% as.character(tpm_data_SorL_stackgenes_low$stage)]
)
fills <- case_when(str_detect(stage_vec, pattern = 'Asex') ~ 'coral1' ,
                   str_detect(stage_vec, pattern = 'Liv') ~ 'plum2',
                   str_detect(stage_vec, pattern = 'Gam') ~ 'lightskyblue1',
                   str_detect(stage_vec, pattern  = 'Bld|Sporo|Ook') ~ 'lightgreen')
k <- 1
for (i in strip_t) {
  j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
  g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
  k <- k+1
}
grid.newpage()
grid.draw(g)
```

```{r table_of_SorL_highest_supptable}
#For a neat table
SorL_kable_index <- rep.int(2, 
                            times = length(unique(tpm_data_SorL_stackgenes$stage)))
names(SorL_kable_index) <- levels(tpm_data_SorL_stackgenes$stage)

tpm_data_SorL_stackgenes_top_table <- tpm_data_SorL_stackgenes %>% 
  mutate(Geneid = stack_group) %>% 
  filter(str_detect(SorL, 
                    pattern = 'ancestral', 
                    negate = TRUE)) %>% 
  group_by(stage) %>% 
  mutate(sum_tpm = sum(total_tpm)) %>% 
  group_by(stage, 
           SorL) %>% 
  slice_max(order_by = total_tpm,
            n = 1) %>% 
  mutate(total_tpm = round(total_tpm, digits = 2),
         prop_tpm = round(total_tpm/sum_tpm, digits = 2)) %>% 
  select(stage, Geneid, SorL, total_tpm, prop_tpm)

tpm_data_SorL_stackgenes_top_table %>% 
  kable() %>% 
  kable_styling() %>% 
  pack_rows(index = SorL_kable_index) %>% 
  scroll_box(width = "100%", height = "500px")
```

## Figure 2B - sub-family proportions 

```{r combined_tpm_stacked_df}

  tpm_data_subfam <- combined_data %>% 
    rownames_to_column(var = 'Geneid') %>% 
    filter(Geneid %in% pir_id) %>% 
    melt(id.vars = 'Geneid', 
         variable.name = 'stage', 
         value.name = 'tpm') %>% 
  mutate(clade = pir_clade$clade[match(Geneid, 
                                       pir_clade$Geneid)]
         ) %>% 
  group_by(clade) %>% 
  mutate(clade = paste0(clade, ' (n = ', n_distinct(Geneid) ,')')) %>% 
    mutate(stack_group = clade) %>% 
    group_by(stage, 
             stack_group,
             clade) %>% 
    dplyr::summarise(total_tpm = sum(tpm)) %>%
  arrange(
          desc(clade),
          total_tpm
)
  

  
tpm_data_subfam <- mutate(tpm_data_subfam,
                   order = row_number())

  # #Apply ordering
  # tpm_data_subfam$stage = factor(tpm_data_subfam$stage, 
  #                         levels = ordering(tpm_data_subfam$stage,
  #                                           order_vector))
  
  #Calculate the range for the y value to make it look nicer
  stacked_max <- tpm_data_subfam %>% 
    ungroup %>% 
    group_by(stage) %>% 
    dplyr::summarise(max_tpm = sum(total_tpm)) %>% 
    .$max_tpm %>% 
    max
  
  max_tpm <- c(
    seq(
      0,
      mround(
        stacked_max,
        500),
      length.out = 11)
  )
```

```{r combined_tpm_stacked_plot}

  stack_label <- 'sub-family'

y <- ggplot(data = tpm_data_subfam, 
            aes(
              x = stage,
              y = total_tpm, 
              fill = clade,
              group = order
            )) +
  geom_bar(
    # position = position_stack(),
    position = 'fill',
    stat = 'identity',
    colour = 'black',
    size = 0.1) +
  scale_fill_manual(
    values = clade_col_num
  ) +
  scale_color_manual('black') + 
  labs(fill = stack_label) +
  scale_y_continuous(breaks = seq(0,1,0.1)) +
  theme_classic() +
  scale_size(range = c(0,3)) +
  ylab(
    bquote('Proportion of '~italic(.('pir'))~' TPM')
  ) + 
  facet_grid(.~stage, 
             scales = "free_x",
             space = "free_x") +
  scale_x_discrete(expand=expand_scale(add=1)) +
  theme(
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x = element_blank(),
    # axis.line.x = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    legend.title = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(3,'mm'),
    legend.key.width = unit(2,'mm'),
    strip.text.x = element_text(size=8, 
                                face = 'bold', 
                                angle = 90),
    strip.background = element_rect(
      color="black", 
      size=0.1, 
      linetype="solid"
    ),
    panel.spacing.x = unit(0, "null")
  ) 

  #This bit is to change the colour of the facet boxes
  
  g <- ggplot_gtable(ggplot_build(y))
  strip_t <- which(grepl('strip-t', g$layout$name))
  stage_vec <- unique(
    
    levels(tpm_data_subfam$stage)
  )
  fills <- case_when(str_detect(stage_vec, pattern = 'Asex') ~ 'coral1' ,
                     str_detect(stage_vec, pattern = 'Liv') ~ 'plum2',
                     str_detect(stage_vec, pattern = 'Gam') ~ 'lightskyblue1',
                     str_detect(stage_vec, pattern  = 'Bld|Sporo|Ook') ~ 'lightgreen')
  k <- 1
  for (i in strip_t) {
    j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
    g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
    k <- k+1
  }
  
  grid.newpage()
  grid.draw(g)
```

## Gametocyte marker gene expression - supp figure 2

```{r}
list_genes <- c(
  list('P28' = 'PBANKA_0514900'),
  list('nek4' = 'PBANKA_0616700'),
  list('map2' = 'PBANKA_0933700'),
  list('hap2' = 'PBANKA_1212600')
)
gam_markers <- unlist(list_genes)

list_tpms_line <- lapply(expt_tpms, 
                         function(input){
                           tpm_data <- input$tpm %>% 
                             filter(Geneid %in% gam_markers) %>% 
                             mutate(Geneid = names(gam_markers)[match(Geneid, gam_markers)]) %>% 
                             melt(id.vars = 'Geneid', variable.name = 'sample', value.name = 'tpm') %>% 
                             mutate(stage = unlist(lapply(strsplit(as.character(sample), 
                                                                   split = '_'),
                                                          `[`,
                                                          1))) %>% 
                             group_by(stage, 
                                      sample,
                                      Geneid) %>% 
                             dplyr::summarise(total_tpm = sum(tpm))
                           
                           #Make stages with no cirs in the sample give a zero 
                           if(!all(colnames(input$tpm) %in% tpm_data$sample)){
                             missing_stages = colnames(input$tpm)[-1][!colnames(input$tpm)[-1] %in% tpm_data$sample]
                             missing_stages_data_zero = data.frame(lapply(missing_stages, function(missing_st){
                               data.frame(stage = unlist(lapply(strsplit(as.character(missing_st), split = '_'),`[`,2)),
                                          sample = missing_st,
                                          total_tpm = 0)
                             }))
                             tpm_data <- bind_rows(tpm_data, missing_stages_data_zero)
                           }
                           
                           #Add experiment column
                           tpm_data$experiment = rep(input$experiment, times = nrow(tpm_data))
                           max_tpm = if(max(tpm_data$total_tpm) < 10){
                             c(seq(0,mround(max(tpm_data$total_tpm),1),1))
                           }else{
                             c(seq(0,mround(max(tpm_data$total_tpm),5),5))
                           }
                           
                           #Get the average data for the line graph
                           tpm_data_line <- tpm_data %>% 
                             ungroup %>% 
                             group_by(stage, 
                                      experiment,
                                      Geneid) %>% 
                             dplyr::summarise(total_tpm_median = median(total_tpm),
                                              Number_of_replicates = n(),
                                              total_tpm_max = max(total_tpm),
                                              total_tpm_min = min(total_tpm)) %>% 
                             arrange(stage)
                           
                           
                           return(tpm_data_line)
                         })

tpm_data_line <- bind_rows(list_tpms_line)

tpm_data_line$stage <- factor(tpm_data_line$stage, 
                             levels = ordering(tpm_data_line$stage,
                                               r_order_vector))

#Calculate the range for the y value to make it look nicer
if(max(
  tpm_data_line$total_tpm_max
) >= 500){
  max_tpm = c(
    seq(
      0,
      mround(
        max(
          tpm_data_line$total_tpm_max
        ),
        500),
      length.out = 11)
  )
}else{
  max_tpm = c(
    seq(
      0,
      mround(
        max(
          tpm_data_line$total_tpm_max
        ),
        10),
      length.out = 11)
  )
}
```

super hacky way to change the scales in each facet from:
https://fishandwhistle.net/post/2018/modifying-facet-scales-in-ggplot2/

```{r}
scale_override <- function(which, scale) {
  if(!is.numeric(which) || (length(which) != 1) || (which %% 1 != 0)) {
    stop("which must be an integer of length 1")
  }
  
  if(is.null(scale$aesthetics) || !any(c("x", "y") %in% scale$aesthetics)) {
    stop("scale must be an x or y position scale")
  }
  
  structure(list(which = which, scale = scale), class = "scale_override")
}

CustomFacetWrap <- ggproto(
  "CustomFacetWrap", FacetWrap,
  init_scales = function(self, layout, x_scale = NULL, y_scale = NULL, params) {
    # make the initial x, y scales list
    scales <- ggproto_parent(FacetWrap, self)$init_scales(layout, x_scale, y_scale, params)
    
    if(is.null(params$scale_overrides)) return(scales)
    
    max_scale_x <- length(scales$x)
    max_scale_y <- length(scales$y)
    
    # ... do some modification of the scales$x and scales$y here based on params$scale_overrides
    for(scale_override in params$scale_overrides) {
      which <- scale_override$which
      scale <- scale_override$scale
      
      if("x" %in% scale$aesthetics) {
        if(!is.null(scales$x)) {
          if(which < 0 || which > max_scale_x) stop("Invalid index of x scale: ", which)
          scales$x[[which]] <- scale$clone()
        }
      } else if("y" %in% scale$aesthetics) {
        if(!is.null(scales$y)) {
          if(which < 0 || which > max_scale_y) stop("Invalid index of y scale: ", which)
          scales$y[[which]] <- scale$clone()
        }
      } else {
        stop("Invalid scale")
      }
    }
    
    # return scales
    scales
  }
)

facet_wrap_custom <- function(..., scale_overrides = NULL) {
  # take advantage of the sanitizing that happens in facet_wrap
  facet_super <- facet_wrap(...)
  
  # sanitize scale overrides
  if(inherits(scale_overrides, "scale_override")) {
    scale_overrides <- list(scale_overrides)
  } else if(!is.list(scale_overrides) || 
            !all(vapply(scale_overrides, inherits, "scale_override", FUN.VALUE = logical(1)))) {
    stop("scale_overrides must be a scale_override object or a list of scale_override objects")
  }
  
  facet_super$params$scale_overrides <- scale_overrides
  
  ggproto(NULL, CustomFacetWrap,
    shrink = facet_super$shrink,
    params = facet_super$params
  )
}

(
  y <- ggplot(data = tpm_data_line, 
              aes(x = stage, 
                  y = total_tpm_median, 
                  fill = experiment, 
                  group = experiment)) +
    geom_bar(position = position_dodge2(preserve = 'single'),
             stat = 'identity') +
    scale_fill_manual(values = c(cbPalette, 'red', 'purple')[1:(length(unique(tpm_data_line$experiment)))]) +
    geom_errorbar(aes(x = stage,
                      ymax = total_tpm_max, 
                      ymin = total_tpm_min),
                  position = position_dodge2(preserve = 'single'),
                  size = 0.25) +
    scale_y_continuous(breaks = max_tpm) +
    theme_classic() +
    theme(axis.title.x=element_blank(),
          axis.text.x=element_text(angle = 90, 
                                   size = 6, 
                                   vjust = 0.5),
          axis.text.y = element_text(size = 5),
          axis.title.y = element_text(size = 8),
          legend.title = element_text(size = 7),
          legend.text = element_text(size = 5),
          legend.key.size = unit(3,'mm'),
          legend.key.width = unit(2,'mm')) +
    scale_size(name = 'Number of genes detected over threshold', 
               range = c(0,3)) +
    ylab(
      bquote('total '~italic(.('pir'))~' TPM')
    ) +
    facet_wrap_custom(~Geneid, 
                      scales = 'free_y',
                      scale_overrides = list(
                        scale_override(3, scale_y_continuous(breaks = seq(0,650,length.out = 11))),
                        scale_override(2, scale_y_continuous(breaks = seq(0,900,length.out = 11))),
                        scale_override(1, scale_y_continuous(breaks = seq(0,310,length.out = 11)))
                      )) +
    annotate("segment", x=-Inf, xend=Inf, y=-Inf, yend=-Inf)
)
#Could also use functions from a package called 'lemon' to add the lines to the top two facets.

ggsave('plots/thesis_chapter3_figure_gametocyteMarkers.png',
       width = 7,
       height = 5,
       units = "in",
       y)
```


## Lower expression of _pir_ genes versus the rest of the genome

```{r low_pir_expression}
set.seed(4908)

gene_expression_df <- combined_data %>% 
    rownames_to_column(var = 'Geneid') %>% 
    melt(id.vars = 'Geneid', 
         variable.name = 'stage', 
         value.name = 'tpm') %>% 
  mutate(pir_gene = ifelse(Geneid %in% pir_id,
                           'pir',
                           'non-pir')) %>% 
  group_by(stage, pir_gene) %>% 
  slice_sample(n = length(pir_id)) %>%
  summarise(mean_tpm = mean(tpm),
            med_tpm = round(median(tpm),
                            digits = 6)) 

g <- ggplot(gene_expression_df,
       aes(x = pir_gene,
           y = log2(1+med_tpm))) +
  xlab('Gene group')+
  ylab('log(Median TPM + 1)')+
  # geom_violin()+
  geom_dotplot(binaxis = 'y',
               stackdir = 'center',
               # binwidth = 1,
               # dotsize = 2,
               method = 'dotdensity') +
  geom_path(aes(group = stage,
                col = stage)) +
  theme_classic() +
  theme(
    legend.text = element_text(size = 9),
    legend.key.size = unit(4, 'mm')
  )

g
```

_pir_ expression on average ranges from minimum mean `r min(gene_expression_df[gene_expression_df$pir_gene == 'pir',]$mean_tpm)` TPM and median `r min(gene_expression_df[gene_expression_df$pir_gene == 'pir',]$med_tpm)` TPM to maximum mean `r max(gene_expression_df[gene_expression_df$pir_gene == 'pir',]$mean_tpm)` TPM and median `r max(gene_expression_df[gene_expression_df$pir_gene == 'pir',]$med_tpm)` TPM. Non-\ _pir_ expression ranges from minimum mean `r min(gene_expression_df[gene_expression_df$pir_gene == 'non-pir',]$mean_tpm)` TPM and median `r min(gene_expression_df[gene_expression_df$pir_gene == 'non-pir',]$med_tpm)` TPM to maximum mean `r max(gene_expression_df[gene_expression_df$pir_gene == 'non-pir',]$mean_tpm)` TPM and median `r max(gene_expression_df[gene_expression_df$pir_gene == 'non-pir',]$med_tpm)` TPM.
