---
title: "tl22-01_VvxSalmonAnalysis"
output: 
  html_document:
    theme: cosmo
    code_download: true
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# Intro

This script analyses the results of a _de novo_ assembly of Plasmodium vivax RNAseq data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      results = TRUE, 
                      message = FALSE 
                      # warning = FALSE
                      )
options(scipen = 999) 
safe_colorblind_palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", 
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")
# Means that numbers are displayed normally not as 1e6 (1000000)
global_save_var <- FALSE
```

```{r loading_packages, include = FALSE, message = FALSE}
# Load the required packages
library('dplyr')
library('readr')
library('stringr')
library('purrr')
library('kableExtra')
library('plotly')
library('rmarkdown')
library('data.table')
library('ggpubr')
library('seqinr')
```

```{r function_ordering}
ordering <- function(to_order, order_vector, fromLast = TRUE){
  order = unlist(sapply(order_vector, 
                        function(x){unique(str_subset(to_order, 
                                                      pattern = paste(x)))}))
  order_missing = unique(to_order[!to_order %in% order])
  unique(c(order, order_missing), fromLast = fromLast)
}
```

```{r function_uniquify}
#Make unique identifiers from a vector of names by appending '_X' where X is the number of the 'replicate'.
uniquify <- function(vector_names, first_one = TRUE){
  new_vector <- c()
  df_num <- data.frame(names = unique(vector_names), rep = 1)
  if(first_one){
    duplicated_names <- unique(vector_names[duplicated(vector_names)])
  }else{
    duplicated_names <- c()
  }
  for(i in 1:length(vector_names)){
    if(vector_names[i] %in% duplicated_names|!first_one){
      name <- paste(vector_names[i],
                    df_num$rep[match(vector_names[i], df_num$names)],
                    sep = '_')
      #Increase rep counter
      df_num$rep[match(vector_names[i], df_num$names)] <- df_num$rep[match(vector_names[i], df_num$names)] + 1
      new_vector[i] <- name
    }else{
      new_vector[i] <- vector_names[i]
    }
  }
  new_vector
}
```

```{r read_quantR}

samplesheet <- read.delim("samplesheet.txt")

#Fix the missing helpful_assembly_name for Kim_U samples
samplesheet$helpful_assembly_name[grepl(samplesheet$library_name, pattern = 'U_{1,2}')] <- rep('Isolate_20', 2)

expt_num.df <- read.csv('expt_codes.csv')

expt_quant <- lapply(unique(samplesheet$Experiment),
                     function(expt){
                       #Get the experiment code.
                       expt_code <- expt_num.df$expt_code[match(expt, 
                                                                expt_num.df$expt)]
                         assembly.list <- lapply(unique(samplesheet$assembly[samplesheet$Experiment == expt]),
                                                 function(assembly_file){
                                                   assembly <- str_extract(assembly_file,
                                                                           pattern = '(?<=cdhitest\\/).+(?=\\/)')
                                                   #Get the file name by pasting the words together
                                                   quant_file <- paste(
                                                     "quantFiles_R",
                                                     expt, assembly, 
                                                     paste0('quant_', 
                                                            assembly,
                                                            '_subfam30pid25lngthI14.csv'
                                                     ),
                                                     sep = '/')
                                                   if(file_test("-f", quant_file)){
                                                     #If the file exists then we can read it in
                                                     quant_csv <- read.csv(quant_file, 
                                                                           header = TRUE)
                                                     return(quant_csv)
                                                   }
                                                 })
                         #Get the names for the list of assemblies, extract the important bits and replace the expt with the expt_code.
                         names(assembly.list) <- str_replace(
                           str_extract(unique(samplesheet$assembly[samplesheet$Experiment == expt]),
                                       pattern = '(?<=cdhitest\\/).+(?=\\/)'), 
                           pattern = expt,
                           replacement = expt_code)
                         return(assembly.list)
                     })

#Change the names of the experiment list to the experiment code.
names(expt_quant) <- expt_num.df$expt_code[match(unique(samplesheet$Experiment), 
                                                 expt_num.df$expt)]

#Remove null elements of list (no quant files)
expt_quant <- expt_quant[!unlist(lapply(expt_quant, 
                                        function(x) is.null(unlist(x))))]

```

# de novo transcript expression

```{r top25}
# Read in the top 25 clusters/sub-family designations
cluster_subfam_25.df <- read.csv(file = 'cluster_subfam_25_df.csv')
```

```{r top_of_each_subfam}
#Get consistent colour values so that each plot will use the same colour scheme.
# Rep each subfamily letter twice and paste '_2' onto the end.
select_subfam <- c('E', 'C_1', 'C_2', 'G', 'I', 'J', 'K_1', 'B_1',
                   'unassigned_1', 'unassigned_2', 'unassigned_3',
                   'other', 'noCluster') %>% ordering(c(.,'noCluster','other'))

cluster_subfam_25.df$subfamSelect <- ifelse(cluster_subfam_25.df$subfam %in% select_subfam, 
                                            cluster_subfam_25.df$subfam,
                                            'other')
cluster_subfam_select.df <- group_by(cluster_subfam_25.df, 
                                     subfamSelect) %>% 
  summarise(total_pirs = sum(total_pirs))

#I want 'none' to be coloured white, and 'other' grey, so I get a palette two colour too short and add them on.
top_subfam_col <- c(rep(safe_colorblind_palette, 
                        length.out = length(select_subfam) - 2), 'white', 'grey')

# top_subfam_col <- c('steelblue1',
#                     'orange', 'firebrick2',
#                     'yellow1',
#                     'springgreen4',
#                     'darkslateblue',
#                     'tomato3',
#                     'grey',
#                     'violetred', 'hotpink1', 'lightsalmon',
#                     'white','cyan3')

top_subfam_col <- c("#006ddb",
                    "#924900", "#db6d00",
                    "#ffff6d",
                    "seagreen3",
                    "#490092",
                    
                    # "#920000",
                    'tomato3',
                    
                    # "#004949",
                    'grey',
                    
                    "#b66dff", "#ffb6db", "#ff6db6",
                    "white", "#009292")

pal <- c("#000000","#004949","#009292","#ff6db6","#ffb6db",
 "#490092","#006ddb","#b66dff","#6db6ff","#b6dbff",
 "#920000","#924900","#db6d00","#24ff24","#ffff6d")
pie(rep(1,length(top_subfam_col)), col = top_subfam_col)
pie(rep(1,length(pal)), col = pal)

names(top_subfam_col) <- select_subfam

# write.csv(data.frame(subfam_assignment = names(top_subfam_col),
#                        colours = unname(top_subfam_col)),
#             file = 'top_subfam_col.csv',
#           row.names = FALSE)
```

How does the number of pir sub-families transcribed in each sample vary by BUSCO score?

```{r pir_diversity_v_BUSCOs}
buscos <- read.table('tl22-01_assembly_info_extra_busco.txt')

subfam_numbers <- lapply(names(expt_quant), 
                         function(expt){
                           #expt <- 'E06'
                           #Doing two graphs, one for all pirs (above threshold) and one with the pir TPM normalised by the number in each sub-family/cluster.
                           expt_subfamNum <- lapply(
                             sort(names(expt_quant[[expt]])),
                             function(assembly){
                               #assembly <- 'E06_A'
                               print(assembly)
                               if(!is.null(expt_quant[[expt]][[assembly]])){
                                 # For this experiment+assembly, melt the pir part of the data frame so that we can calculate per sample the number of sub-families transcribed above 1 TPM.
                                 expt_quant[[expt]][[assembly]] %>% 
                                   filter(pir) %>% 
                                   reshape2::melt(.,
                                                  id.vars = c('pir','Name', 'subfam'),
                                                  measure.vars = str_subset(colnames(.), 
                                                                            pattern = 'TPM'),
                                                  value.name = 'tpm') %>% 
                                   group_by(variable, 
                                            subfam) %>% 
                                   summarise(max_tpm = max(tpm)) %>% 
                                   #Summarise should drop 'subfam' from grouping.
                                   filter(max_tpm >= 1) %>% 
                                   summarise(num_subfam = n_distinct(subfam)) %>% 
                                   mutate(expt = expt,
                                          assembly_name = assembly)
                               }
                             }) %>% 
                             bind_rows
                         }) %>% 
  bind_rows
#Warning "no non-missing arguments to max; returning -Inf" may indicate that there aren't any pirs present.
buscos <- mutate(rowwise(buscos), assembly_name = str_replace(assembly_name, experiment, experiment_num))
#Fix the messed up Gunalan names again
subfam_numbers <- mutate(subfam_numbers, assembly_name = str_replace(assembly_name, 
                                                                     'Guna19',
                                                                     'E07'))
busco_subfamNum.df <- inner_join(subfam_numbers, buscos, by = 'assembly_name')

expt_colors <- unique(buscos$expt_colour)
names(expt_colors) <- unique(buscos$experiment_num)
expt_colors <- expt_colors[sort(names(expt_colors))]

ggplot(busco_subfamNum.df,
       aes(x = perc_busco_plasmo_nonMissing,
           y = num_subfam,
           col = expt)) +
  geom_point(size = 5, alpha = .75) +
  theme_classic() +
  scale_color_manual(values = expt_colors[names(expt_colors) %in% busco_subfamNum.df$expt])

```

```{r barchart}

#Make samplesheet column for nicer sample names using other information
samplesheet_samplename <- 
  mutate(samplesheet, 
         Expt_code = expt_num.df$expt_code[match(Experiment, 
                                                 expt_num.df$expt)]) %>% 
  rowwise() %>% 
  mutate(sample_name = paste(c(Expt_code, 
                               Samples.combined, 
                               Sample_stage),
                             collapse = '_'),
         assembly_exptCode = str_replace_all(assembly, 
                                         pattern = Experiment, 
                                         replacement = Expt_code))

samplesheet_samplename$sample_name <- uniquify(samplesheet_samplename$sample_name, 
                                               first_one = TRUE) 
#Which type of bar chart am I making?
geom_col_pos <- 'fill'
# geom_col_pos <- 'stack'
           
# Make a stacked bar chart of the
lapply(names(expt_quant),
       function(expt){
         #expt <- 'E06'
         #Doing two graphs, one for all pirs (above threshold) and one with the pir TPM normalised by the number in each sub-family/cluster.
         subfampir.list <- lapply(
           c('totalTPM','absolute_subfam','normalised_subfam'),
           function(absolute_or_norm){
             #absolute_or_norm <- 'absolute_subfam'
             experiment_plot <- lapply(
               sort(names(expt_quant[[expt]])),
               function(assembly){
                 #assembly <- 'E06_A'
                 if(!is.null(expt_quant[[expt]][[assembly]])){
                   
                   helpful_assembly_name <- unique(samplesheet_samplename$helpful_assembly_name[
                     str_detect(samplesheet_samplename$assembly_exptCode ,
                                assembly)
                   ])
                   
                   #Set a size adjustment for when there are loads of graphs together and the axes need to be smaller
                   size_adjust <- ifelse(length(names(expt_quant[[expt]])) > 6,
                                         0.75,
                                         1)
                   
                   subfam_top.df <- expt_quant[[expt]][[assembly]] %>%
                     filter(pir) %>%
                     #Include only the subfams in the top20
                     mutate(subfam_top = ifelse(subfam %in% select_subfam,
                                                subfam,
                                                'other')) %>%
                     reshape2::melt(.,
                                    id.vars = c('subfam_top','pir','Name'),
                                    measure.vars = str_subset(colnames(.),
                                                              pattern = 'TPM'),
                                    value.name = 'tpm') %>%
                     #Filtering step for TPM
                     filter(tpm >= 1) %>%
                     rowwise() %>%
                     mutate(sample_name = samplesheet_samplename$helpful_sample_name[
                       str_detect(samplesheet_samplename$prefix,
                                  str_remove(variable,
                                             pattern = '_TPM'))
                     ]) %>%
                     #Arranging by TPM here means that when factorised the transcripts will be ordered by tpm value
                     arrange(sample_name,subfam_top, tpm)
                   
                   if(absolute_or_norm == 'normalised_subfam'){
                     #Normalise by the total number of pirs in each cluster
                     subfam_top.df <- mutate(subfam_top.df,
                                             number = cluster_subfam_select.df$total_pirs[
                                               match(subfam_top,
                                                     cluster_subfam_select.df$subfamSelect)
                                             ],
                                             tpm = tpm/cluster_subfam_select.df$total_pirs[
                                               match(subfam_top,
                                                     cluster_subfam_select.df$subfamSelect)
                                             ])
                   }
                   
                   #If there are more rows than zero (i.e. there are pirs)
                   if(dim(subfam_top.df)[1] > 0){
                     plot <- ggplot(data = subfam_top.df,
                                    aes(x = factor(sample_name,
                                                   levels = ordering(sample_name,
                                                                     order_vector = c("^6hr", "before"))),
                                        y = tpm))
                     
                     #For the 'totalTPM' graph I want the bar charts to be 'stacked' so the genes are added on top of each other and we can see the summed TPM values for each sample.
                     if(absolute_or_norm == 'totalTPM'){
                       geom_col_pos <- 'stack'
                       plot <- plot + geom_col(
                         position = geom_col_pos,
                         show.legend = FALSE
                       )
                     }else{
                       # For the graphs with the sub-families shown I want the bar charts to be 'fill' so that I'm visualising the ratios of the sub-family members.
                       geom_col_pos <- 'fill'
                       # colour and size so that each gene block has a small outline
                       plot <- plot + geom_col(
                         colour = 'black',
                         size = 0.02,
                         position = geom_col_pos,
                         show.legend = FALSE,
                         aes(fill = factor(subfam_top, 
                                           levels = names(top_subfam_col)))
                       )
                     }
                     
                     plot <- plot + xlab('') + ylab('TPM') +
                       
                       # facet_wrap(~subfam_top, ncol = 2, scales = "free_y") +
                       #Make scale_fill manual values depend on varying length of the palette so it can accomodate having more than 12 variables
                       scale_fill_manual(values = top_subfam_col) +
                       theme_classic() +
                       #Going to keep the axis text only if it's of interest, or the first plot. Must be in if{}else{} because ifelse breaks it.
                       theme(axis.text.x = element_text( size = 11*size_adjust),
                             axis.title.y = element_blank(),
                             axis.text.y = element_text(size = 15*size_adjust),
                             plot.title = element_text(size = 7*size_adjust)) +
                       ggtitle(paste0(expt,'_',assembly,'_',helpful_assembly_name))
                     
                     if(geom_col_pos != 'fill'){
                       #If not using 'fill' then need a nicer y axis
                       plot <- plot +
                         scale_y_continuous(breaks = signif(
                           seq(0,
                               max(
                                 group_by(subfam_top.df,
                                          variable) %>%
                                   summarise(total_tpm = sum(tpm)) %>%
                                   .$total_tpm
                               ),
                               length.out = 5),
                           digits = 2)
                         )
                     }
                     
                     return(plot)
                   } else {
                     return(NULL)
                   }
                   
                 }
               })
           })
         
         names(subfampir.list) <- c('totalTPM','absolute_subfam','normalised_subfam')
         
         if(!is.null(unlist(subfampir.list))){
           # If list returned isn't just NULL
           
           #Go into the 'all_pirs' data to get the indexes for splitting the graphs into two groups.
           # This is to make the graphs look neater if there's more than ten datasets.
           # Need to do this on the 'all_pirs' data so that the two sets of graphs will be directly comparable. Doing this for
           #   each in turn may lead to different indexes due to different numbers of 'subfam_pirs'.
           
           #Get index of NULL values to remove. 
           notNull.index <- !unlist(lapply(subfampir.list[[1]],
                                           function(plot) is.null(plot)))
           #Remove the NULL elements of lists
           subfampir.list <- lapply(subfampir.list, function(x.list) x.list[notNull.index])
           assembly_names.notNull <- sort(names(expt_quant[[expt]]))[notNull.index]
           
           plot.list <- lapply(
             c("absolute_subfam",
               "normalised_subfam"),
             function(absolute_or_norm){
               twoplot.list <- lapply(
                 seq_along(subfampir.list[[1]]), 
                 function(subfampir.index){ 
                   apl <- subfampir.list[['totalTPM']][subfampir.index][[1]]
                   bpl <- subfampir.list[[absolute_or_norm]][subfampir.index][[1]]
                   assembly <- assembly_names.notNull[subfampir.index]
                   helpful_name <- unique(samplesheet_samplename$helpful_assembly_name[
                     str_detect(samplesheet_samplename$assembly_exptCode ,
                                assembly)
                   ])
                   
                   
                   if(global_save_var){
                     ggsave(
                       plot = cowplot::plot_grid(apl, bpl, ncol = 2),
                       filename = paste0(
                         'plots/salmonQuant_plots/tl22-01_',
                         expt, '_', absolute_or_norm, '_', assembly,
                         '_quantSalmon_stacked_subfam30pid25lngthI14.pdf'
                       ),
                       width = 7,
                       height = 5,
                       units = "in",
                       dpi = 330
                     )
                   }
                   
                   return(cowplot::plot_grid(apl, bpl, ncol = 2))
                 }
               )
             }
           )
         }
       })

```

Notes:
* Cheng19 samples particularly divergent?
* Some samples far more heterogeneous than others.
* Strain specific differences show overwhelming inclusion of some families over others.
** e.g. A few Kim19 show no other assembled sub-family beyond C_1.
** Kim17 isolate 2
* E_1s and C_1s tend to dominate expression (E_1 and C_1 are the largest groups).
* Most pir expression comes from those which cannot be straightforwardly characterised by existing sub-families.

* Kim samples mostly show major loss of pir transcription before and after chloroquine admin, although could this be explained by other factors? Also not seen for every sample.
* Muller19 Thai shows that the two sporozoite pirs are expressed, albeit at low quantities, neither in a subfam.

* Overall the proportions of sub-fams are very consistent! 
** Could this be an artefact of clustering and choosing the largest clusters to plot?

```{r legend}
#Just showing the legend without a plot to accompany the above.
legend.plot <- ggplot(data.frame(subfam = names(top_subfam_col), top_subfam_col), 
       aes(subfam, fill = subfam)) + 
  geom_bar() +
  scale_fill_manual(values = top_subfam_col) +
  guides(fill = guide_legend(title = 'Subfamily designation',
                             ncol = 1)) +
  guides(fill = guide_legend(override.aes = list(alpha=1)))+ # showing the point in the legend
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.5, 0.5), # move the legend to the center
        legend.title = element_blank(),
        legend.key = element_rect(fill='NA'),
        panel.grid = element_blank(),
        panel.border = element_rect(colour = "black", fill='white', size=1))

if(global_save_var){
  ggsave(
    plot = legend.plot,
    filename = 'plots/salmonQuant_plots/tl22-01_legend_quantSalmon_stacked_subfam30pid25lngthI14.pdf',
    width = 7,
    height = 5,
    units = "in",
    dpi = 330
  )
}
```


```{r save_barchart_data}
expt_barchart.list <- lapply(names(expt_quant),
                           function(expt){
                             #expt <- 'E01'
                             #Doing two graphs, one for all pirs (above threshold) and one with the pir TPM normalised by the number in each sub-family/cluster.
                             subfampir.list <- 
                               lapply(sort(names(expt_quant[[expt]])),
                                      function(assembly){
                                        #assembly <- 'E01_SMRU1'
                                        experiment_plot <- if(!is.null(expt_quant[[expt]][[assembly]])){
                                          
                                          helpful_assembly_name <- unique(samplesheet_samplename$helpful_assembly_name[
                                            str_detect(samplesheet_samplename$assembly_exptCode ,
                                                       assembly)
                                          ])
                                          
                                          #Set a size adjustment for when there are loads of graphs together and the axes need to be smaller
                                          size_adjust <- ifelse(length(names(expt_quant[[expt]])) > 6,
                                                                0.75,
                                                                1)
                                          
                                          subfam_top.df <- expt_quant[[expt]][[assembly]] %>%
                                            filter(pir) %>%
                                            #Include only the subfams in the top20
                                            mutate(subfam_top = ifelse(subfam %in% select_subfam,
                                                                       subfam,
                                                                       'other')) %>%
                                            reshape2::melt(.,
                                                           id.vars = c('subfam_top','pir','Name'),
                                                           measure.vars = str_subset(colnames(.),
                                                                                     pattern = 'TPM'),
                                                           value.name = 'tpm',
                                                           variable.name = 'sample') %>%
                                            #Filtering step for TPM
                                            filter(tpm >= 1) %>%
                                            rowwise() %>%
                                            mutate(sample_name = samplesheet_samplename$helpful_sample_name[
                                              str_detect(samplesheet_samplename$prefix,
                                                         str_remove(sample,
                                                                    pattern = '_TPM'))
                                            ],
                                            assembly = assembly,
                                            expt = expt) %>%
                                            #Arranging by TPM here means that when factorised the transcripts will be ordered by tpm value
                                            arrange(sample_name,subfam_top, tpm)
                                          
                                          return(subfam_top.df)
                                        } 
                                      })
                             names(subfampir.list) <- sort(names(expt_quant[[expt]]))
                             
                             subfampir.df <- bind_rows(subfampir.list[!unlist(lapply(subfampir.list, 
                                                                                     plyr::empty))])
                             return(subfampir.df)
                           })

names(expt_barchart.list) <- names(expt_quant)

expt_barchart.list <- expt_barchart.list[!unlist(lapply(expt_barchart.list,
                                                        plyr::empty))]

expt_barchart.df <- bind_rows(expt_barchart.list)

```

```{r highest_prop_subfam}

lapply(expt_barchart.list, 
       function(expt.df){
         expt.df %>% 
           group_by(subfam_top, 
                    assembly,
                    sample,
                    sample_name) %>%
           summarise(tpm_sum = sum(tpm))          %>%
           group_by(assembly, sample, sample_name) %>%
           slice_max(order_by = tpm_sum, n = 10)
       })

```

```{r normalised}
lapply(expt_barchart.list, 
       function(expt.df){
         subfam_top.df <- mutate(expt.df,
                                 number = cluster_subfam_25.df$total_pirs[
                                   match(subfam_top,
                                         cluster_subfam_25.df$subfam)
                                 ],
                                 tpm_norm = tpm/cluster_subfam_25.df$total_pirs[
                                   match(subfam_top,
                                         cluster_subfam_25.df$subfam)
                                 ]) %>% 
           group_by(subfam_top, 
                    assembly,
                    sample,
                    sample_name) %>%
           summarise(tpmNorm = sum(tpm_norm), 
                     tpm = sum(tpm),
                     num = n_distinct(Name))          %>%
           group_by(assembly, sample, sample_name) %>%
           slice_max(order_by = tpmNorm, n = 5)
       })
#other is NA right now - need to fix this.
```

Unassigned_3 seems to be enriched for transcription given its size. 

Sub-families E and C_2 are large groups but still higher transcribers.

```{r heatmap_clustering}
heatmap.df <- lapply(names(expt_quant),
       function(expt){
         #expt <- 'E01'
           lapply(sort(names(expt_quant[[expt]])),
                  function(assembly){
                    #assembly <- 'E01_SMRU1'
                    
                    #Filter for ext=pressed pirs
                    pir_tpmover1.df <- expt_quant[[expt]][[assembly]] %>% 
                      filter(pir) %>% rowwise %>% filter(
                             across(matches('TPM'), ~ any(.x > 1))) 
                    
                    if(nrow(pir_tpmover1.df) == 0){
                      # print('no pirs')
                      return()
                    }
                    
                    helpful_assembly_name <- unique(samplesheet_samplename$helpful_assembly_name[
                      str_detect(samplesheet_samplename$assembly_exptCode ,
                                 assembly)
                    ])
                    
                      #Calculate the TPM of each sub-family as a proportion of pir transcription
                      reshape2::melt(pir_tpmover1.df,
                                     id.vars = c('pir','Name', 'subfam'),
                                     measure.vars = str_subset(colnames(pir_tpmover1.df), 
                                                               pattern = 'TPM'),
                                     value.name = 'tpm',
                                     variable.name = 'sample') %>% 
                      group_by(sample, subfam) %>% 
                      summarise(subfam_tot = sum(tpm)) %>% 
                      mutate(subfam_prop = subfam_tot/sum(subfam_tot)*100) %>% 
                      mutate(sum_prop = sum(subfam_prop),
                             sample_name = paste0(expt, '_',
                                                  helpful_assembly_name, 
                                                  '_',
                                                  samplesheet_samplename$helpful_sample_name[
                                                    str_detect(samplesheet_samplename$prefix,
                                                               str_remove(sample,
                                                                          pattern = '_TPM'))
                                                  ])) %>% 
                      dcast(sample_name ~ subfam, value.var = 'subfam_prop') 
                  })
       }) %>% bind_rows

heatmap.df <- tibble::column_to_rownames(heatmap.df, 
                                 var = 'sample_name')
heatmap.df <- t(as.matrix(heatmap.df))

library('ComplexHeatmap')
heatmap.df[is.na(heatmap.df)] <- 0
Heatmap(heatmap.df, 
        cluster_rows = TRUE, 
        cluster_columns = TRUE,
        column_split = 10)

```

```{r node-df_tpmvalues}

# Add the TPM values to the node.df dataset alongside a threshold expressed/not expressed

# Get a list of all the de novo transcripts and their median TPM. 
#  Shouldn't be any with the same name unless by exceptional chance?
gene_quant <- lapply(names(expt_quant), 
                     function(expt.name) {
                       expt.list <- expt_quant[[expt.name]]
                       # print(names(expt.list))
                       lapply(names(expt.list), 
                              function(quant.name) {
                                quant.df <- expt.list[[quant.name]]
                                if(!is.null(quant.df)){
                                  #Melt the dataframe to calculate the median TPM value for each de novo transcript across all samples.
                                  quant.df %>% 
                                    reshape2::melt(.,
                                                   id.vars = c('pir','Name'),
                                                   measure.vars = str_subset(colnames(.), 
                                                                             pattern = 'TPM'),
                                                   value.name = 'tpm') %>% 
                                    group_by(Name, pir) %>% 
                                    summarise(med_tpm = median(tpm),
                                              max_tpm = max(tpm)) %>% 
                                    mutate(label_no_end = paste(quant.name, Name, sep = '_'),
                                           expt_assembly = paste(expt.name, quant.name, sep = '_'))
                                }
                              }) %>% 
                         bind_rows
                     }) %>% 
  bind_rows

# Need to add the assembly name to the transcript name (as in node.df) because some are the same, which I find surprising!
# Some are from the same assemblies but not all of them:
# gene_quant[gene_quant$Name == 'NODE_24950_length_245_cov_2.19474',]

```

```{r over1tpm_listFiles}
#Note that node_tpm.df includes none-quantified genes, like P01/Sal1/Zhu16 and the Rangel genes.
# Will use over or equal to 1 TPM in at least one sample (max_tpm > 1) as the threshold for active transcription.
expressed_pirTranscripts <- filter(gene_quant, pir, max_tpm>=1)

paste0('Number of transcribed pirs = ',
       nrow(expressed_pirTranscripts),
       ' and number of non-transcribed pirs = ',
       nrow(filter(gene_quant, pir)) - nrow(expressed_pirTranscripts))

G18G19_hmmerVir200aa.fa <- read.fasta('G18_B20_C19_Z16_M19_R20_K1719_G19-cat_scspades_0.95cdhitest_hmmerPfam-PlasmodiumVir_200aa.fasta')

names_no_end <- str_remove(names(G18G19_hmmerVir200aa.fa), pattern = '_[[:digit:]]{1,2}$')

#These aren't all unique, I guess this doesn't impact anything later.

#Subset the fasta file by the pirs which are actively transcribed.
# Note that this may be greater than expected due to the fact that there may be multiple pir ORFs for one expressed pir transcript.
# Also it is smaller than the number of expressed_pirTranscripts because not all of the expressed_pirTranscripts have a >200aa ORF.
# Note that the pirs are defined as such based on the *_scspades_0.95cdhitest_hmmerPfam-PlasmodiumVir.list files (so before 200aa filtering).

G18G19_hmmerVir200aa_tpmOver1.fa <- G18G19_hmmerVir200aa.fa[
  names(G18G19_hmmerVir200aa.fa)[names_no_end %in% expressed_pirTranscripts$label_no_end]
  ]
print(paste0('Number of ORFs (>200aa ORF tpm over 1) = ',
      length(G18G19_hmmerVir200aa_tpmOver1.fa)))
print(paste0('Number of unique transcripts (>200aa ORF tpm over 1) = ',
      length(unique(str_extract(names(G18G19_hmmerVir200aa_tpmOver1.fa), 
                   pattern = '.+(?=_[[:digit:]]$)')))))

table(str_extract(names(G18G19_hmmerVir200aa_tpmOver1.fa), pattern ='.+(?=_NODE)'))

# write.fasta(G18G19_hmmerVir200aa_tpmOver1.fa,
#             names(G18G19_hmmerVir200aa_tpmOver1.fa),
#             file.out = 'G18_B20_C19_Z16_M19_R20_K1719_G19-cat_scspades_0.95cdhitest_hmmerPfam-PlasmodiumVir_orf200aa_tpmOver1.fasta')

```

```{r Num-pir-transcript_v_Num-transcript-total}
assembly_info.df <- read.delim('tl22-01_assembly_info.txt') %>% 
  filter(!grepl(assembly, pattern = '0\\.9c|0\\.99'),
         !grepl(assembly, pattern = 'SRX5158513_SRR8347514|SRX5158512_SRR8347515'))  
#Filter out any none-0.95 assembly made. And potential Kim19_A duplicate?

#Order experiments by years
expt_num.vec <-ordering(unique(assembly_info.df$experiment), 
                                   order_vector = c(16:22))
#Get experiment numbers and add '0' to single digit numbers.
names(expt_num.vec) <-  paste0('E', 
                              sapply(seq_along(expt_num.vec), function(x){
                                ifelse(nchar(x)==1, 
                                       paste0('0', x),
                                       x)
                              }))

#Manually add Gunalan19 and _Boonkaew20 (in addition to Guna19 and Boonkaew20)
expt_num.vec <- c(expt_num.vec, E07 = 'Gunalan19', E08 = '_Boonkaew20')

#Get the relevant info from samplesheet otherwise loads of rows will be duplicated when joining.
samplesheet_small.df <- mutate(samplesheet, assembly_file = assembly) %>% 
  select(assembly_file, Cat.Sample.stage.s.) %>% unique

assembly_info_samplesheet.df <- left_join(assembly_info.df, samplesheet_small.df, by = 'assembly_file')

assembly_info_samplesheet.df <- mutate(assembly_info_samplesheet.df,
                                        experiment_num = names(expt_num.vec)[match(experiment,
                                                                                  expt_num.vec)])

#Get consistent shapes and colours for the graphs

shapes_val <- 1:length(unique(assembly_info_samplesheet.df$Cat.Sample.stage.s.))
names(shapes_val) <- unique(assembly_info_samplesheet.df$Cat.Sample.stage.s.)

colours_val <- safe_colorblind_palette[1:length(unique(names(expt_num.vec)))]
names(colours_val) <- unique(names(expt_num.vec))

expt_num.df <- data.frame(expt = expt_num.vec, 
                          expt_code = names(expt_num.vec), 
                          col = c(colours_val,colours_val['E07'],colours_val['E08']))
# write.csv(expt_num.df,
#           file = 'expt_codes.csv',
#           row.names = FALSE)

  
( plot <- ggplot(assembly_info_samplesheet.df, 
       aes(x = num_transcripts,
           y = num_uniq_mRNA_ORFs_over200aaLength,
           col = experiment_num,
           shape = Cat.Sample.stage.s.
       )) +
  geom_point(size = 1, stroke = 2, alpha = .75) +
  scale_shape_manual(values=shapes_val,
                     name = 'Life cycle stage',
                     guide = guide_legend(ncol = 2)) +
  scale_color_manual(values = colours_val,
                     name = 'Experiment',
                     guide = guide_legend(ncol = 2)) +
  xlab('Number of assembled transcripts') +
  ylab(expression(paste('Number of putative ', italic('pir'), ' transcripts'))) +
  theme_classic() )

if(global_save_var){
  ggsave(plot = plot,
         'plots/assemblies_numPirTranscripts_totalTranscripts.pdf',
         device = 'pdf',
         width = 7,
         height = 5,
         units = "in")
}
```

ISSUE: Why is the number of sequences in 'G18_B20_C19_Z16_M19_R20_K1719_G19-cat_scspades_0.95cdhitest_hmmerPfam-PlasmodiumVir_200aa.fasta' larger than sum(assembly_info_samplesheet.df£$num_uniq_mRNA_pirHMMerORFs_nomatch95known_over200aa)

Observations:
* Some drop-outs between the samples? Perhaps because one transcript was exclusively assembled from one sample.
